Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-22 Manuscript/manuscript.Rmd
cleanup

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect genomic regulation [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Approximately 60-80% of TADs remain stable across cell types [@Schmitt:2016aa], suggestingthat the boundaries of TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @flavahan2016insulator] [??? REF; !!!added references]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

Identifying the mechanisms underlying the formation of TADs are a complex and active area of research. The distinct patterns of different proteins and functional elements point toward the opportunity of using computational approaches in identifying which epigenomic features are most predictive of the TAD boundaries. [??? Better to mention those that came up in your analysis, otherwise, it is unnecessary information; !!!editted paragraph] Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries [@mourad2016computational; @hong2017computational; @huang2015predicting] [??? Do not understand this sentence; !!!editted sentence]. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. Moreover, many methods ignore key characteristics of domain data, such as heavily imbalanced number ot boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, this will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions mostly constitute as resampling techniques, in which random sampling is used to either replicate or removed data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data] [??? REFS; !!!added references]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study we focus our attention toward data level solutions so that all results may be classifier independent, and thus, more easily comparable.

@@

Replicate Hi-C data for the GM12878, K562, ... cell lines were downloaded from GEO GSE63525 (Experiments HIC001 to HIC018, HIC069 to HIC 074, ... respectively; produced using the MbolI restriction enzyme) [@rao20143d]. The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained.

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance (in bases) from the center of the bin to the center of the nearest region of a respective genomic annotation region was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figure 1)

@@

In order to evaluate the problem of imbalanced classes, we then evaluated 4 re-sampling techniques (in addition to no re-sampling).

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE. We chose the parameters that would create perfectly balanced classes. The `DMwR` package in R was used to perform SMOTE (version 0.4.1).

5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernel density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` package [??? Is it correct? Same version as DMwR?; !!!fixed] in R used to perform ROSE (version 0.0.3).

@@

A classification algorithm in the form of a random forest was performed on the various resampled and reduced training sets and validated using the same test [??? Training?; !!!clarified; different training sets due to the different resampling schemes, but validated using the same holdout set] set for all models in the ensemble. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

>>>>

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect genomic regulation [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Approximately 60-80% of TADs remain stable across cell types [@Schmitt:2016aa], suggestingthat the boundaries of TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @flavahan2016insulator]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

Identifying the mechanisms underlying the formation of TADs are a complex and active area of research. The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes [??? REFs] prompt the use of computational approaches to identify which epigenomic features are most predictive of the TAD boundaries. Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries [@mourad2016computational; @hong2017computational; @huang2015predicting]. Although several useful insights have been made, such as the convergent binding of CTCF [??? Elaborate with refs.], many methods ignore key characteristics of domain data, such as heavily imbalanced number of boundaries vs. other regions. In a classification setting [??? are we talking about classification or prediction?], when one is attempting to discriminate between regions with and without TAD boundaries, this will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions mostly constitute as resampling techniques, in which random sampling is used to either replicate or removed data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study we focus our attention toward data level solutions so that all results may be classifier independent, and thus, more easily comparable.

@@

Replicate Hi-C data for the GM12878, K562, ... cell lines were downloaded from GEO GSE63525 (Experiments HIC001 to HIC018, HIC069 to HIC 074, ... respectively; produced using the MbolI restriction enzyme) [@rao20143d]. [??? Need supplementary ref.] The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained.

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance (in bases) from the center of the bin to the center of the nearest region of a respective genomic annotation region [??? Decide whether to use "annotation" or "feature"] was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figure 1).

@@

In order to evaluate the problem of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling).

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE. We chose the parameters that would create perfectly balanced classes. The `DMwR` R package was used to perform SMOTE (version 0.4.1).

5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernel density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` R package was used to perform ROSE (version 0.0.3).

@@

A classification algorithm in the form of a random forest was performed on the various resampled and reduced training sets and validated using the same test set for all models in the ensemble. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.
